\input{./preamble.tex}
\usepackage{quiver}
\usepackage{listings}
\usepackage{fancyvrb,xcolor}
\usepackage{tcolorbox}
\usepackage{tabto}
\usepackage{wrapfig}
\definecolor{grey}{rgb}{0.4,0.4,0.4}
\definecolor{graay}{rgb}{0.8,0.8,0.8}
\definecolor{dg}{rgb}{0.1,0.4,0.1}
\definecolor{skybl}{rgb}{0.85,0.92,0.95}
\newcommand{\catname}[1]{{\normalfont\textbf{#1}}}
\newcommand{\Set}{\catname{\sffamily Set }}
\newcommand{\Ring}{\catname{\sffamily Ring }}
\newcommand{\Group}{\catname{\sffamily Group }}
%\newcommand{\inner}[2]{
%\left<#1,#2\right>
%}
\newcommand{\pb}[2]{
\begin{tcolorbox}[
width=\textwidth,
colback={skybl},
title={#1},
%outer arc=0mm,
%colbacktitle=dg,
%coltitle=black,
%colupper=white
]  
#2
\end{tcolorbox} 
}
\newcommand{\w}{\wedge}
%\usepackage{ifpdf,mla}
%\graphicspath{ {./} }
\title{\vspace{-1in}\textbf{Algebra Qualifying Exams}\\
{\large{Rutgers - the State University of New Jersey}}\\
{\small{\href{https://www.math.rutgers.edu/docman-lister/math-main/academics/graduate/qualifying-exam/1266-syl2014/file}{Syllabus}
}}}


\author{\vspace{-2in}Nilava Metya}
\date{\vspace{-0.2in}}

\begin{document}
\maketitle
\tableofcontents
\newpage



\section*{\href{https://www.math.rutgers.edu/docman-lister/math-main/2630-wq-jan-2023-all-3/file}{Spring 2023}}
\addcontentsline{toc}{section}{Spring 2023}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Groups}{ 
Classify all groups of order 309, up to isomorphism.
%\import{./soln/S23/}{1.tex}
}




%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Groups}{ 
Let $A$ be the abelian group with generators $x,y,z$ and the relations
$$4x+3y+z=0, x+2y+3z=0, 3x+2y+5z=0$$ Show that $A$ is a cyclic abelian group, and determine its order.
%\import{./soln/S23/}{2.tex}
}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Linear Algebra}{ 
Let $A$ be a complex $n \times n$ matrix. Prove that there is an invertible complex $n \times n$ matrix $B$ such that $AB = BA^t$. ($A^t$ is the transpose of $A$.)
\import{./soln/S23/}{3.tex}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Rings}{ 
Prove that the subring $\Z[3i]$ of $\C$ is not a Principal Ideal Domain.
%\import{./soln/S23/}{4.tex}
}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Rings}{ 
If $R = \Z[x]$, show that the sequence $R\stackrel{f}{\to} R^2\stackrel{g}{\to} R$ is exact, where $ f(a) = (ax, -2a)$ and $g(c,d) = 2c+dx$.
%\import{./soln/S23/}{5.tex}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\section*{\href{https://www.math.rutgers.edu/docman-lister/math-main/2577-august-wq-exams-combined/file}{Fall 2022}}
\addcontentsline{toc}{section}{Fall 2022}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Groups}{ 
Let $G$ be a finite simple group. Prove that $G \times G$ has exactly $4$ normal subgroups (including $G \times G$) if and only if $G$ is non-abelian.
%\import{./soln/F22/}{1.tex}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Rings}{ 
Let $R$ be a principal ideal domain and $I,J$ be ideals of $R$. Show that
$I \cap J = IJ$ holds if and only if $I =0$ or $J = 0$ or $I + J = R$.
%\import{./soln/F22/}{2.tex}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Linear Algebra}{ 
Let $A \in M_n(\R)$ be a symmetric matrix with real coefficients. Show that all eigenvalues of $A$ are non-negative if and only if $A = P^T P$ for some matrix $P \in M_n(\R)$.
.\import{./soln/F22/}{3.tex}
}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Rings}{ 
Let $R$ be an integral domain and $R[x,y,z]$ the polynomial ring in three variables over $R$. Show that $I = \left<x^3 ,y^2, y^3 -z^2y\right> \subseteq R[x, y, z]$ is a prime ideal. \\
Hint: Show that $I$ is the kernel of a ring homomorphism $R[x, y, z] \to R[t]$.
%\import{./soln/F22/}{4.tex}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Linear Algebra}{ 
Let $A$ and $B$ be commuting complex matrices. Assume that $B \notin \C[A]$, that is, $B$ cannot be written as a polynomial in $A$. Show that some eigenspace of $A$ has dimension at least two.
%\import{./soln/F22/}{5.tex}
}








%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\section*{\href{https://www.math.rutgers.edu/docman-lister/math-main/2515-spring-2022-written-exams/file}{Spring 2022}}
\addcontentsline{toc}{section}{Spring 2022}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Rings}{ 
Prove that the rings $\Q[x]/(x^2 - 1)$ and $\Q \oplus \Q$ are isomorphic.
%%\import{./soln/S22/}{1.text}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Groups}{ 
Let $p$ be a prime. Show that any element of order $p$ in 
$\GL_2(\Z/p\Z)$ can be conjugated to the matrix $\begin{bmatrix}
1 & 1\\
0 & 1
\end{bmatrix}$.
%\import{./soln/S22/}{2.text}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Fields}{ 
Let $a$ and $b$ be elements of a field of order $2^n$ where $n$ is odd. Prove that if $a^2 +ab+b^2 =0$ then $a=b=0$.
\import{./soln/S22/}{3.text}
}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Linear Algebra}{ 
Let $A, B$ be linear operators on a nonzero finite-dimensional vector space $V$ over $\C$ such that $A^2 = B^2 = \id$. Prove that there exists a nonzero subspace $W$ of $V$ which is invariant under $A$ and $B$ and $\dim W \leq 2$.
\import{./soln/S22/}{4.text}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Linear Algebra}{ 
Let $A$ be a complex $n \times n$ matrix. Let $a_k$ denote the dimension of the null space of $A^k$ (in particular, $a_0 = 0$). Prove that $a_k + a_{k+2} \leq 2a_{k+1}$ for all $k \geq 0$.
%\import{./soln/S22/}{5.text}
}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\section*{\href{https://www.math.rutgers.edu/docman-lister/math-main/2454-wq-august-2021/file}{Fall 2021}}
\addcontentsline{toc}{section}{Fall 2021}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Groups}{ 
Let $G$ be a group and $Z(G)$ the center of $G$. Show that the group $G/Z(G)$ does not have prime order. Find a group $G$ such that $G/Z(G)$ has $4$ elements.
%\import{./soln/F21/}{1.tex}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Rings}{ 
Show that every prime ideal $P$ in $\Z[x]$ which is not principal contains a prime number.
%\import{./soln/F21/}{2.tex}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Groups}{ 
Show that every finite noncyclic group is a finite union of proper subgroups, and that if a group maps surjectively to a finite noncyclic group then it is a finite union of proper subgroups and use this to determine for which positive integers the product of $n$ copies of the integers is a finite union of proper subgroups.
%\import{./soln/F21/}{3.tex}
}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Linear Algebra}{ 
Let $A$ and $B$ be two square matrices over a field $F$. Suppose $\diag(A, A)$ and $\diag(B, B)$ are similar. Show that $A$ and $B$ are similar.
%\import{./soln/F21/}{4.tex}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Groups}{
\begin{enumerate}[label=(\alph*), leftmargin=*,noitemsep]
\item Suppose that $p$ and $q$ are distinct primes and a group $G$ is generated by elements of order $p$ and also by elements of order $q$. Show that any homomorphism of $G$ to an abelian group is trivial.
\item Show that for $n \geq 5$ the alternating group $A_n$ of even permutations of $n$ objects is generated by elements of order $2$, and also by elements of order $3$, so that for such $n$ the only homomorphisms to abelian groups are trivial.
\end{enumerate}
%\import{./soln/F21/}{5.tex}
}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\section*{\href{https://www.math.rutgers.edu/docman-lister/math-main/2419-wq-j21-alg-compvar-realvar-website/file}{Spring 2021}}
\addcontentsline{toc}{section}{Spring 2021}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Rings}{ 
The following are four classes of commutative rings, in alphabetical order:
\begin{itemize}[noitemsep]
\item fields
\item integral domains
\item principal integral domains
\item unique factorization domains
\end{itemize}
These are contained in one-another, in some order, so that
$A_1\subsetneq A_2\subsetneq A_3\subsetneq A_4$.\begin{enumerate}[label=(\alph*),noitemsep]
\item Determine the order.
\item Give an example in each class to show that the inclusions are proper.
\end{enumerate}
\import{./soln/S21/}{1.tex}
}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Rings}{ 
\begin{enumerate}[label=(\alph*),noitemsep]
\item  If $R$ is a commutative ring, define what it means for $R$ to be Noetherian and state Hilbert’s basis theorem.
\item Give an example of a non-Noetherian commutative ring.
\end{enumerate}
\import{./soln/S21/}{2.tex}
}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Groups}{ 
Let $G$ be a group of order $105$ and let $P_3$, $P_5$, and $P_7$ be Sylow $3$, $5$, and $7$ subgroups, respectively. Assuming the Sylow theorems, prove
the following:
\begin{enumerate}[label=(\alph*),noitemsep]
\item At least one of $P_5$ or $P_7$ is normal in $G$.
\item $G$ has a cyclic subgroup of order $35$.
\item Both $P_5$ and $P_7$ are normal in $G$.
\end{enumerate}
%\import{./soln/S21/}{3.tex}
}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Linear Algebra}{ 
Find all similarity classes of $2\times2$ matrices $A$ with entries in $\Q$ satisfying $A^4 = I$. What are the corresponding rational canonical forms?
%\import{./soln/S21/}{4.tex}
}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Linear Algebra}{ 
\begin{enumerate}[label=(\alph*),noitemsep]
\item Find the possible Jordan Canonical Forms of any matrix such that $A^4 = I$ over $F = \mathbb F_5$.
\item  Give an example of a matrix $B$ over $F = \mathbb F_3$ satisfying $B^4 = I$, such that $B$ is not diagonalizable.
\end{enumerate}
%\import{./soln/S21/}{5.tex}
}







%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\section*{\href{https://www.math.rutgers.edu/docman-lister/math-main/2375-august-2020-algebra-complex-analysis-real-analysis/file}{Fall 2020}}
\addcontentsline{toc}{section}{Fall 2020}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Linear Algebra}{ 
Prove that for any pair of commuting $n \times n-$matrices with complex entries there exists a common eigenvector.
%\import{./soln/F20/}{1.tex}
}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Groups}{ 
Prove that there exists no simple group of order $56$.
%\import{./soln/F20/}{2.tex}
}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Rings}{ 
Prove that a ring which contains a principal ideal ring $R$, and which is contained in the field of fractions of $R$, is a principal ideal ring.
%\import{./soln/F20/}{3.tex}
}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Linear Algebra}{ 
Let $A$ and $B$ be two projection linear maps in a vector space over a field $K$. Prove that if $A + B$ is a projection linear map and $\text{char}K\neq 2$ then $AB = BA = 0$.
\import{./soln/F20/}{4.text}
}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Groups}{ 
Prove that in the group $\Q/\Z$ for any natural number $n$ there exists exactly one subgroup of order $n$.
%\import{./soln/F20/}{5.tex}
}








%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\section*{\href{https://www.math.rutgers.edu/docman-lister/math-main/2327-jan20writtenexams/file}{Spring 2020}}
\addcontentsline{toc}{section}{Spring 2020}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Algebra}{ 
Suppose that $A$ is a not necessarily commutative, finite dimensional associative algebra with a unit over a field $F$ and $P \unlhd A$ is a two-sided ideal such that for $a,b\in A, ab\in P \implies a\in P$ or $bP\in P$. Show that $A/P$ must be a division algebra (i.e. every nonzero element has a multiplicative inverse).
%\import{./soln/S20/}{1.tex}
}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Groups}{ 
Show that every group of order $2020$ contains a unique (and hence normal) subgroup of order $505$.
%\import{./soln/S20/}{2.tex}
}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Linear Algebra}{ 
Let $M$ be a matrix with integer entries.
\begin{enumerate}[label=(\alph*)]
\item Prove that the minimal polynomial of $M$ over $\C$ $$f_{\min}(t) = t^k + \sum_{i=0}^{k-1}a_it^i$$ has integer coefficients.
\item Prove that if $M$ is diagonalizable over $\Q$ then there exists an integer $N$ such that the matrix $M \mod p$ is diagonalizable over $\Z/p\Z$ for all $p > N$.
\end{enumerate}
%\import{./soln/S20/}{3.tex}
}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Rings}{ 
Let $F$ be a field and let $L$ be the ring of Laurent polynomials $L = F[x,x^{-1}]$ (it is the subring of $F(x)$ generated over $F$ by $x$ and $x^{-1}$). We consider $L$ as a module over the ring of polynomials $R = F[x]$.
(a) Show that $L$ is not a finitely generated module over $R$.
(b) Show that every finitely generated submodule of $L$ is free with a single generator.

%\import{./soln/S20/}{4.tex}
}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\rule{\textwidth}{2pt}
\pb{Rings}{ 
Let $R$ be a commutative integral domain and let $I \unlhd R$ be an ideal.
\begin{enumerate}[label=(\alph*)]
\item Show that every alternating bilinear form $$f:I\times I\to R$$ is zero.
\item Show that if $R$ is a principal ideal domain, then every alternating bilinear form $f: I\times I\to M$ to any $R-$module $M$ is zero.
\end{enumerate}
%\import{./soln/S20/}{5.tex}
}







\end{document}